---
title: 'Linking stormwater control performance to stream ecosystem outcomes: incorporating a performance metric into effective imperviousness'
subtitle: 'Table, figures and text requiring data extraction'
author: 'Christopher J. Walsh, Matthew J. Burns, Tim D. Fletcher, Darren G. Bos, Peter Poelsma, Joshphar Kunapo and Sam J. Imberger'
date: 'School of Ecosystem and Forest Sciences, The University of Melbourne, 500 Yarra Boulevard, Burnley Victoria 3121, Australia '
output: 
  word_document:
    reference_docx: officedown_template.docx
csl: plos.csl
bibliography: references.bib
editor_options: 
  chunk_output_type: inline
---

```{r loadData, include = FALSE, warning=FALSE, message=FALSE}
requiredPackages <- c("RPostgreSQL","sf","lubridate","mcr","RColorBrewer",
                      "httr","rjson","scales","dplyr","here","RPostgreSQL",
                      "data.table","RMySQL")
lapply(requiredPackages, require, character.only = TRUE)
source(here("code","download.OSF.file.R"))
source(here("code","BACRIfunctions.R"))  
knitr::opts_chunk$set(echo = FALSE, dev = "pdf")
#Check all gpkg files have been downloaded to the data file, and download any that are missing.
source(here("load_ld_scms_tables.R"))
dpi <- 600
```

Table 1. Summary statistics for stormwater control in the 6 experimental sub-catchments.  Stormwater control measure (SCM) projects were commissioned as single projects and consisted of one or more SCMs. Note that the counts for D8 include those listed for its sub-catchment D4, and those for L4 include those for its sub-catchments L1, Ls and Ln. Treated EIA is the effective impervious area draining to one or more SCMs.  The estimates of runoff volume harvested or lost, and volume filtered, are modelled volumes given the arrangement of SCMs in each catchment at the end of the study period assuming rainfall pattern of the average year that was used to calculate S (1965-1966).  
```{r echo = FALSE}
ei_fin <- rbind(tail(ei_101$iats,1),tail(ei_103$iats,1),tail(ei_53$iats,1),
                tail(ei_36$iats,1),tail(ei_74$iats,1),tail(ei_71$iats,1))
cat_stats <- data.frame(pipeID = c(101,103,53,36,74,71),
                        site = final_stats$site,
                        cat = c("D4","D8","Ln","Ls","L1","L4"),
                        n_scms = NA,
                        n_scm_projects = NA,
                        eia_s1_ha = round(1e-4 * final_stats$EI * ei_fin$carea,1),
                        ia_treated_ha = round(1e-4 * ei_fin$carea * (final_stats$EI - final_stats$EI_s),1),
                        ia_treated_perc = round(100 * (final_stats$EI - final_stats$EI_s) / final_stats$EI,1),
                        vol_reduced_ML = round((final_stats$V_u - final_stats$V_m) * 1e-6),
                        vol_reduced_perc = round(100* (final_stats$V_u - final_stats$V_m) / final_stats$V_u,1),
                        vol_filtered_L = round(final_stats$F_m * 1e-6),
                        vol_filtered_perc = round(100* (final_stats$F_m) / final_stats$V_u,1),
                        stringsAsFactors = FALSE)
SCMs$subc <- subcs$trib[match(SCMs$pipeID,subcs$pipeID)]
scmProjects$subc <- subcs$trib[match(scmProjects$pipeID,subcs$pipeID)]
for(i in 1:6){
  allSCMs <- SCMs$scmID[SCMs$subc == cat_stats$site[i]]
  allSCMs <- allSCMs[!allSCMs %in% scmsDecommissioned$scmID]
  cat_stats$n_scms[i] <- length(allSCMs)
  cat_stats$n_scm_projects[i] <- length(scmProjects$projectID[scmProjects$subc == cat_stats$site[i]])
}
cat_stats$n_scms[cat_stats$cat == "D8"] <- sum(cat_stats$n_scms[cat_stats$cat %in% c("D4","D8")])
cat_stats$n_scm_projects[cat_stats$cat == "D8"] <- sum(cat_stats$n_scm_projects[cat_stats$cat %in% c("D4","D8")])
cat_stats$n_scms[cat_stats$cat == "L4"] <- sum(cat_stats$n_scms[cat_stats$cat %in% c("L4","L1","Ln","Ls")])
cat_stats$n_scm_projects[cat_stats$cat == "L4"] <- sum(cat_stats$n_scm_projects[cat_stats$cat %in% c("L4","L1","Ln","Ls")])

cat_stats_t <- data.frame(t(cat_stats[,4:12]))
names(cat_stats_t) <- cat_stats[,3]
statistic_col <- c("Number of SCMs", "Number of SCM projects","EIA (ha)",
                   "Treated EIA (ha)", "% EIA treated", "Runoff volume harvested/lost (ML/y)",
                   "% Runoff volume harvested/lost", "Runoff volume filtered (ML/y)",
                   "% Runoff volume filtered")
cat_stats_t <- data.frame(`Stormwater control` = statistic_col,
                        cat_stats_t, stringsAsFactors = FALSE)
#remove % stats, which are presented in Fig. 3
cat_stats_t <- cat_stats_t[!grepl("%", statistic_col),] 
ft <- table_for_word(cat_stats_t, font = "Times New Roman", pgwidth = 6.69)
ft <- flextable::colformat_num(ft, i = 1:2, digits = 0)
ft <- flextable::align_nottext_col(ft)
ft
```
##### page break  

### Text from discussion requiring data from the database  

```{r}
#Number of target properties with SCMs
#First in L4.
n_target_props_L4 <- sum(parcels$pipeID < 100 & parcels$parcelType == "property" & parcels$roofAreaCon > 0)
# #Some properties in L4 split across subcatchments - can only find these in database by address
# addresses <- sqlQuery("SELECT DISTINCT address, \"parcelID\" FROM parcels WHERE \"pipeID\" < 100 AND \"parcelType\" = 'property' AND \"roofAreaCon\" > 0", 
#                       "lsc_dbs_scms")
# dim(addresses[grep("_", addresses$address),])  #28 - so remove these from n_target_props
n_target_props_L4 <- n_target_props_L4 - 28
n_hits_L4 <- length(unique(scmProjects$parcelID[scmProjects$round %in% c(1:4,"SO")]))
L4_prop_target_props <- n_hits_L4/n_target_props_L4

#Second in D8: No split properties D8
n_target_props_D8 <- sum(parcels$pipeID >= 100 & parcels$parcelType == "property" & parcels$roofAreaCon > 0)
#Systems were installed on some disconnected properties: exclude these
scmProjects$parcelType <- parcels$parcelType[match(scmProjects$parcelID, parcels$parcelID)]
scmProjects$roofAreaCon <- parcels$roofAreaCon[match(scmProjects$parcelID, parcels$parcelID)]
n_hits_D8 <- length(unique(scmProjects$parcelID[scmProjects$round %in% c("DBS tanks round 1","DBS tank round 2" ) & scmProjects$parcelType == "property" & scmProjects$roofAreaCon > 0]))
D8_prop_target_props <- n_hits_D8/n_target_props_D8
```

Despite several campaigns of community engagement, which overcame many barriers to community participation, SCMs were installed on only 28% of the private properties with drainage connection in the L4 catchment  [@bos_brown_2015]: this number grew to `r round(100*L4_prop_target_props)`% in the following 4 years, because of a planning provision placed on the catchment [@gov_vic_2019].  Two engagement rounds in the D8 catchment by Melbourne Water achieved SCMs on `r round(100*D8_prop_target_props)`% of the connected properties.  

```{r}
load("data/scmProjects_EB.rda")
# SCMs installed as part of the experiment in L4 and D8.
lsc_eb <- scmProjects_EB[scmProjects_EB$pipeID < 100 & 
                         scmProjects_EB$round %in% c("D",1:4,"SO","PN","PL"),]
dbs_eb <- scmProjects_EB[scmProjects_EB$pipeID >= 100 & 
                        scmProjects_EB$round %in% c("DBS tanks round 1",
                                                    "DBS tanks round 2","DBS KCC raingardens"),]
petrol_tank_specs <- tanks[grep("Petrol",tanks$projectID), ]
# Two tanks - RPLT536 overflows to RPLT537 (and leaks via a filter to stormwater)
petrol_term_tank <- "RPLT537"
#The petrol station has its own pipeID - no other impervious surfaces.
petrol_pipeID <- scmProjects$pipeID[grepl("Petrol",scmProjects$projectID)]
petrol_budget <- budget_scm_on_datex(petrol_term_tank,"2018-01-01")
# Note the above budget includes the leak from RPLT536, as the budget considers
# fate of all water falling upstream of the scm - but it is only being used to 
# calculate the total usage, which was all drawn from RPLT537
petrol_eb <- EB_subc_on_datex(petrol_pipeID,"2018-01-01")
petrol_s <- 1 - petrol_eb$scm_stats$EB/petrol_eb$scm_stats$EB_max
petrol_imp_ha <- round(petrol_eb$scm_stats$EB_max/100,2) # 0.33 ha
petrol_annual_demand_L <- sum(eval(parse(text = petrol_tank_specs$other[petrol_tank_specs$scmID == "RPLT537"])))
petrol_perc_runoff_harvested <- 100 * (1 - petrol_eb$scm_stats$V_m/petrol_eb$scm_stats$V_u)
petrol_perc_demand_met <- 100*sum(petrol_budget$budget$use)/petrol_annual_demand_L
petrol_perc_Ls_carea <- 100 * petrol_imp_ha*1e4/(tail(ei_ts$ei[ei_ts$sitecode == "LSS0001"],1)*cats$carea_m2[cats$sitecode == "LSS0001"])
morrisons_at_end <- data_on_datex(8,"2019-12-01")
morrisons_budget <- budget_scm_on_datex(scmProjects_EB$term_scm[scmProjects_EB$projectID == 'PL-8-Pembroke (Morrisons Reserve)'],"2019-12-01")
morrisons_annual_demand_L <- sum(c(1062629.8,850103.8,354209.9,283367.9,0.0,0,0,0,0.0,0.0,177105.0,814682.8)) +
  550*18.9*250  #irrigation demand plus toilet flushing demand in schools...
morrisons_perc_demand_met <- round(100*sum(morrisons_budget$budget$use)/morrisons_annual_demand_L,0)
perc_Ln_ia_morrisons_ia <- round(100 *sum(ia$area_m2[ia$pipeID == 8 & ia$conn == 1])/sum(ia$area_m2[ia$subc == "LSN0001" & ia$conn == 1]))
morrisons_stats <- scmProjects_EB[scmProjects_EB$projectID == 'PL-8-Pembroke (Morrisons Reserve)',]
morrisons_perc_runoff_captured <- round(100*with(morrisons_stats, (V_u - V_m)/V_u))
l4_roofs_untreated <- round(100*(1 - sum(parcels$roofAreaCon[match(lsc_eb$parcelID[!is.na(lsc_eb$parcelID)], parcels$parcelID)])/
sum(parcels$roofAreaCon[parcels$pipeID < 100])))
l4_perc_pave <- round(100*sum((parcels$paveAreaCon)[parcels$pipeID < 100])/
                      sum((parcels$paveAreaCon + parcels$roofAreaCon)[parcels$pipeID < 100]))
```

The SCMs that were installed as part of the experiment varied widely in performance (range of *S* `r paste(round(range(lsc_eb$S),2), collapse = "-")` in L4, and `r paste(round(range(dbs_eb$S),2), collapse = "-")` in D4), underlining the utility of *EI~S~* for ensuring a comparable measure of experimental effect within our study, and with other studies. The temporal and inter-site patterns of $\Delta$*EI~S~* (Fig. 5B), weighted by SCM performance, differed substantially from $\Delta$*EI~S0~* (Fig. 5C) treating all SCMs uniformly like @roy_etal_2014 and @loperfido_etal_2014.  Optimal performance of individual SCMs, as measured by *S*, required a combination of harvesting and loss of at least 55% of impervious runoff (Fig. 3D), and retaining and filtering of 19-45% (Fig. 3C) to achieve a low frequency of uncontrolled flow to the stream. Some infiltration systems performed poorly for *EI~SF~* because they filtered too much water (`r sum(lsc_eb$F_m > lsc_eb$F_pasture)` systems in L4, `r sum(dbs_eb$F_m > dbs_eb$F_pasture)` in D8). However, at the catchment scale, infiltration volumes were lower than optimal because of the lack of SCM coverage, and the local excess of infiltration from these SCMs improved performance at the catchment-scale.  However, to achieve near-full SCM coverage without exceeding infiltration targets, substantially greater harvesting and loss of impervious runoff than we achieved is require to avoid more widespread infiltration excess. 

Water demand was enough to meet the volume reduction target of 55% in `r sum(lsc_eb$perc_runoff_lost >= 55)` SCM projects in L4 and `r sum(dbs_eb$perc_runoff_lost >= 55)` in D8. All of these projects harvested runoff for domestic uses from relatively small areas: roofs of < 625 m^2^ (80% < 270 m^2^).  Two projects harvested water for large demands from larger areas, but were of limited influence at the catchment scale.   Tanks and treatment systems receiving runoff from the entire `r petrol_imp_ha` ha of impervious area of the catchment's petrol station provide `r round(petrol_perc_demand_met)`% of its car wash’s `r round(petrol_annual_demand_L*1e-6,1)` ML/y demand.  Despite this large demand, the harvested water only accounted for `r round(petrol_perc_runoff_harvested)`% of the runoff generated by the station's impervious area, and *S* for this project was `r round(petrol_s,2)`. The influence of this high-performing project was ultimately small, as the petrol station's impervious area constituted only `r round(petrol_perc_Ls_carea,1)`% of the effective impervious surfaces in the Ls catchment.  The largest-scale harvesting scheme collected runoff from a large educational precinct in the Ln catchment with `r round(scmProjects_EB$EB_max[grep("Morr",scmProjects_EB$projectID)]/100, 1)` ha of impervious surfaces providing `r morrisons_perc_demand_met`% of the toilet-flushing demands of the precinct's schools, and the dry-weather irrigation demand of its playing fields, which totalled ~`r round(morrisons_annual_demand_L * 1e-6)` ML.  However, this large harvesting volume, comprised only `r morrisons_perc_runoff_captured`% of the runoff generated by its impervious catchment. Despite, some SCM projects achieving the volume reduction target, generally roof runoff on private property’s produced more water than the property’s demand. Furthermore, the lack of SCM coverage meant no runoff from `r l4_roofs_untreated`% of the L4 catchment’s roofs was harvested, and very little from paved surfaces and roads, which make up about `r l4_perc_pave`% of the catchments’ impervious area.   

##### page break  


### Figure captions  

Fig. 1.  Examples of stream ecosystem response variables as a function of effective imperviousness (EI).  
(A) Median filterable reactive phosphorus (FRP) concentrations and (B) SIGNAL score (a biotic index). Piecewise regressions [solid black line with 95% confidence intervals in the grey polygon as derived by @walsh_etal_2005a] are compared to linear regressions against log$_{10}$(EI + 0.1), solid red line with dotted 95% confidence intervals. Five overlapping points in (A) are shown by slight jittering.

Fig. 2.  The 11 study sites considered in this paper and their catchments.
Sites and catchment boundaries are coloured by their status as control, experimental or reference. The 7 primary catchments of the original experiment are italicized. Impervious areas in each catchment are shaded by their connection to the stormwater drainage system in 2019. (Impervious areas outside catchments are not shown.)  Inset maps show the location of the main map (grey rectangle) in Victoria, Australia. All data and code to produce this figure are available at https://osf.io/57azq/.

Fig. 3.  Stormwater control spread and effectiveness in each of the 6 experimental sub-catchments. (A) Number of SCM projects per ha of effective impervious area (EIA). (B) Percentage of EIA that drains to at least one SCM. (C) Percentage of impervious runoff harvested or lost to evapo-transpiration in SCMs. (D) Percentage of impervious runoff filtered through SCMs or to soil. In (C) and (D), the pink polygons indicate the target percentages proposed by @walsh_etal_2012. D8 and L4 bars are shaded to indicate that these sub-catchments contain other sub-catchments. D4 is part of D8, and L1, Ln and Ls are part of L4: see Fig 1.  

Fig. 4. Time series of effective imperviousness (EI) variants achieved in each of the experimental catchments.  Subscripts denote variants. S1: all effective impervious areas upstream of stormwater control measures (SCMs) unweighted (assuming SCMs have no effect, S = 1). S0: all effective impervious areas upstream of SCMs set to 0 (assuming all SCMs retain stormwater perfectly). For other subscripts, all effective impervious surfaces upstream of SCMs are weighted by the stormwater stream impact metric (S), or one of its sub-indices, of the most downstream SCM in any train of SCMs. SF: the filtered-flow sub-metric. SR: the runoff frequency sub-metric. SV: the volume reduction sub-metric. SW: the water quality sub-metric.

Fig. 5. Time series of predictors of effects of urban stormwater runoff and control in the 11 study catchments.
(A) *degrd*: log~10~(x + 0.1)-transformed effective imperviousness assuming no stormwater control (*EI~S1~*).  (B) *restr*, or $\Delta$*EI~S~*, the change in log~10~(x + 0.1)-transformed EI weighted by the stormwater stream impact metric, S. (C) $\Delta$*EI~S0~*, the  change in log~10~(x + 0.1)-transformed EI with all impervious surfaces draining to an SCM weighted zero) for the two control (C) sub-catchments, three reference (R) sub-catchments and the six experimental sub-catchments, indicated by code (see Fig 2). Note all control and reference values in panels B and C equal 0.

Fig 6. Results from the hierarchical linear model of two simulated response variables to urban stormwater runoff and control.
(A, B) show results for *y*, simulated to be unaffected by stormwater control (as indicated by *restr*), while (C, D) show results for *y1*, simulated to be equally affected by stormwater runoff (as indicated by *degrd*, as it was at the start of the study) and stormwater control.  (A, C)  show coefficient estimates for each effect, showing that the model correcly estimated these effects. (The *degrd* effect on both variables was simulated to increase over time, resulting in the mean *degrd* effect for *y1* being greater than the *restr* effect).  (B, D) show model predictions under four treatment scenarios at the end of the study: as they would have been without any stormwater control (red), the *restr* effect ($\Delta$*EI~S~* reduction) actually achieved (orange), a change in $\Delta$*EI~S~* of -1 (i.e. a 10-fold reduction in EIS, light green), and reducing EIS to zero (reference condition, dark green).  Each estimate is illustrated by the median (point), 80th (thick error bars) and 95th (thin error bars) percentile credible intervals.


##### page break

```{r fig1_frp_v_ei, warning=FALSE, message=FALSE, fig.width = 7.48, fig.height = 3.5}
#pdf("images/Fig 1.pdf", width = 7.48, height = 3.5)
# dpi <- 600
# tiff("images/Fig 1.tiff", width = 7.48*dpi, height = 3.58*dpi, res = 600, compression = "lzw")
layout(matrix(c(1,2,3,3), 2, 2, byrow = TRUE), heights = c(10,1), widths = c(10,10))
par(mar = c(2,4,1,1))
#data used by Walsh et al 2005 taken from "SUSE NABS issue/FRP stats/data.txt"
# and "SUSE NABS issue/EPTE/data.txt"
 lFRP50 <- c(-1.56864E+00, -2.22185E+00, -2.09691E+00, -2.22185E+00, -2.52288E+00, -1.88606E+00, -2.04576E+00, -2.52288E+00, -2.52288E+00, -2.69897E+00, -2.00000E+00, -2.39794E+00, -1.92082E+00, -1.92082E+00)
 EI <- c(1.95167E-01, 4.26416E-03, 3.62570E-02, 4.19733E-03, 0.00000E+00, 9.52280E-02, 4.63966E-01, 0.00000E+00, 0.00000E+00, 1.16397E-03, 1.03733E-02, 6.65377E-03, 3.82980E-01, 5.47130E-02)
 #jitter to reveal overlapping points - first element of multiplier is axis range
 jittery <- c(0,-0.05,0,0.05,-0.1,0,0,0,0.1,0,0,0,0,0) * 1.5 * 0.05
 jitterx <- c(0,0,0,0,-0.05,0,0,0.05,-0.05,0,0,0,0,0) * 2.6 * 0.05
 xgon <- seq(0,0.4,0.001)
 y <- -2.6 + 57*xgon
 y[y > -2] <- -2
 plot(log10(xgon +0.001),y, typ = 'n',
      xlim = c(-3,-0.39), ylim = c(-3,-1.5), axes = FALSE,
      xlab = "", ylab = "")
 yupper <- -2.3 + 490*xgon
 yupper[yupper > -1.8] <- -1.8
 ylower <- -2.8 + 13*xgon
 ylower[ylower > -2.1] <- -2.1
 polygon(c(log10(xgon +0.001),rev(log10(xgon +0.001))),
         c(yupper,rev(ylower)), col = gray(0.75), border = gray(0.75))
 lines(log10(xgon + 0.001),y)
 axis(1, at = c(-3.5,log10(c(0,0.003,0.01,0.03,0.1,0.3)+0.001)),
                labels = c("",c(0,0.003,0.01,0.03,0.1,0.3)*100))
 axis(2, at = log10(c(0.0001,0.001,0.003,0.01,0.03,0.1,0.3,1,3,10,30)),
      lab = c("",0.001,0.003,0.01,0.03,0.1,0.3,"1.0","3.0",10,30), las = 1)
 title(ylab = "Median FRP (mg/L)")
 points(log10(EI +0.001) + jitterx,
        lFRP50 + jittery, pch = 21, bg = gray(0.5))
 lY <- log10(EI +0.001)
 linmod <- lm(lFRP50 ~ lY)
 newdata <-  data.frame(lY = log10(c(0.0001,0.001,0.003,0.01,0.03,0.1,0.3,1,3,10,40)))
 abline(linmod, lty = 1, col = "red")
 linmodCLs <- predict(linmod, newdata, interval = "predict")
# note this tutorial says that the default interval = predict returns the 95% confidence limits.
 matplot(as.vector(newdata["lY"]),linmodCLs[,2:3], type = 'l', col = "red", lty = 3, add = TRUE)
  title(main = "A.", adj = 0, line = 0, cex.main = 0.9)

 sigE=c(5.29000E+00, 4.32000E+00, 6.73000E+00, 5.61000E+00, 6.35000E+00, 6.79000E+00, 4.72000E+00, 4.82000E+00, 6.47000E+00, 6.42000E+00, 6.71000E+00, 6.12000E+00, 6.88000E+00, 4.00000E+00, 5.05000E+00) 
EI_sig <- c(3.87607E-02, 1.95167E-01, 4.26416E-03, 3.62570E-02, 4.19733E-03, 0.00000E+00, 9.52280E-02, 4.63966E-01, 0.00000E+00, 0.00000E+00, 1.16397E-03, 1.03733E-02, 6.65377E-03, 3.82980E-01, 5.47130E-02)
 
 plot(log10(EI_sig +0.001),sigE, typ = 'n',
      xlim = c(-3,-0.39), ylim = c(3,8), axes = FALSE,
      xlab = "", ylab = "")
 axis(1, at = c(-3.5,log10(c(0,0.003,0.01,0.03,0.1,0.3)+0.001)),
                labels = c("",c(0,0.003,0.01,0.03,0.1,0.3)*100))
 axis(2, at = 1:9, las = 1)
 title(ylab = "SIGNAL score")

 y <- 6.7 - 31*xgon
 y[y < 4.5] <- 4.5
 yupper <- 6.9 - 21*xgon
 yupper[yupper < 4.7] <- 4.7
 ylower <- 6.5 - 40*xgon
 ylower[ylower < 4.2] <- 4.2
 polygon(c(log10(xgon +0.001),rev(log10(xgon +0.001))),
         c(yupper,rev(ylower)), col = gray(0.75), border = gray(0.75))
 lines(log10(xgon + 0.001),y)
 
 points(log10(EI_sig +0.001),
        sigE, pch = 21, bg = gray(0.5))
 lY <- log10(EI_sig +0.001)
 linmod <- lm(sigE ~ lY)
 newdata <-  data.frame(lY = log10(c(0.0001,0.001,0.003,0.01,0.03,0.1,0.3,1,3,10,40)))
 abline(linmod, lty = 1, col = "red")
 linmodCLs <- predict(linmod, newdata, interval = "predict")
 matplot(as.vector(newdata["lY"]),linmodCLs[,2:3], type = 'l', col = "red", lty = 3, add = TRUE)
 title(main = "B.", adj = 0, line = 0, cex.main = 0.9)
 par(mar = c(0,0,0,0))
 plot.new()
  title(xlab = "             Effective imperviousness (%)", line = -1.5)
#dev.off()
```
  
Fig. 1. 

```{r fig2_sitemap, warning=FALSE, message=FALSE, fig.width=6.8, fig.height=8.75, dev = "png",dpi = 300}
#tiff("images/Fig 2.tiff", width = 6.8*dpi, height = 8.75*dpi, res = 600, compression = "lzw")
lo <- layout(matrix(c(1,2,1,3,1,4,1,5,1,6),5,2,byrow=TRUE),widths = c(5.75,1.25), 
       heights = c(1.25,1.25,4,1.25,1.25))
par(mai = c(0.028,0.028,0.028,0.028)) 
plot(catMap11$geometry, lty = 3, lwd = 0.5, border = "white")
#censored to exclude IA newer than October 2009 (i.e. 2009 IA)
catIA_2009 <- catIA[is.na(catIA$constructionDate) | (!is.na(catIA$constructionDate) &
                catIA$constructionDate < "2009-10-01"),]
plot(catIA_2009$geometry,
     col = c(gray(0.6),gray(0.85),gray(0.85))[match(catIA_2009$conn,c(1,0,0.5))],
     border = c(gray(0.6),gray(0.85),gray(0.85))[match(catIA_2009$conn,c(1,0,0.5))],
     lwd = 0.5, add = TRUE)
#col converts conn = 0.5 to 0 (those surfaces that would be connected if the pits weren't full)..
#only applies in Wicks subcatchment
plot(streams$geometry, col = "dodgerblue", lwd = 2, add = TRUE)
plot(catMap11$geometry, lty = 3, lwd = 2, border = catMap11$col, add = TRUE)
plot(siteMap11$geometry, pch = 21, bg = siteMap11$col, cex = 1.1, add = TRUE)
siteLabels$font[siteLabels$font == 1] <- 2 #make both normal and italic fonts bold
text(st_coordinates(siteLabels),siteLabels$sites_p, cex = 1, font = siteLabels$font)
arrows(st_coordinates(siteLabels)[siteLabels$siteLabel == "LS"][1]-200,
       st_coordinates(siteLabels)[siteLabels$siteLabel == "LS"][2],
       st_coordinates(siteMap11)[siteMap11$sitecode == "LSS0001"][1],
       st_coordinates(siteMap11)[siteMap11$sitecode == "LSS0001"][2], length = 0.05)
arrows(st_coordinates(siteLabels)[siteLabels$siteLabel == "LM"][1],
       st_coordinates(siteLabels)[siteLabels$siteLabel == "LM"][2]+100,
       st_coordinates(siteMap11)[siteMap11$sitecode == "LIS0001"][1],
       st_coordinates(siteMap11)[siteMap11$sitecode == "LIS0001"][2], length = 0.05)
arrows(st_coordinates(siteLabels)[siteLabels$siteLabel == "LN"][1],
       st_coordinates(siteLabels)[siteLabels$siteLabel == "LN"][2]-100,
       st_coordinates(siteMap11)[siteMap11$sitecode == "LSN0001"][1],
       st_coordinates(siteMap11)[siteMap11$sitecode == "LSN0001"][2], length = 0.05)
raster::scalebar(2000, xy = c(355500,5801500), label= "2 km")
rng <- par("usr")
rngdat <- data.frame(x1 = rng[1], x2 = rng[2], y1 = rng[3], y2 = rng[4])
box()
plot.new()
legend("center", lty = c(3,3,3), lwd = c(2,2,2), 
       col = c(RColorBrewer::brewer.pal(3,"Dark2")), title = "Catchments",
       legend = c("Control","Experimental","Reference"), text.col = "white")
legend("center", pch = 21, pt.bg = c(RColorBrewer::brewer.pal(3,"Dark2")),
       title = "Catchments",
       legend = c("Reference","Experimental","Control"), bty = 'n')
plot.new()
legend("center", 
       pch = 15, pt.cex = 1., col = c(gray(0.6),gray(0.85)),
       legend = c("Connected","Unconnected"), title = "Impervious areas")
rngPoly <- st_sfc(st_polygon(list(rbind(rng[c(1,3)], rng[c(1,4)], rng[c(2,4)], 
                                 rng[c(2,3)],rng[c(1,3)]))))
rngPoly <- st_sf(a = 1, geom = rngPoly)
st_crs(rngPoly) <- 28355
rngPoly <- rngPoly %>% st_transform(4283)
par(mai = c(0.028,0.028,0.028,0.028))
plot.new()
plot(Australia_GDA94_GCS$geometry)
plot(Victoria_GDA94_GCS$geometry, col = "white", add = TRUE)
polygon(c(143.5,146.5,146.5,143.5),c(-36.5,-36.5,-39.1,-39.1), col = scales::alpha(gray(0.5),0.5))
plot(rngPoly$geom, col = "black", add = TRUE)
title(main = "Australia", adj = 0, font.main = 1, line = -1)
box()
bounds <- st_sfc(st_point(c(143.5,-36.5)), st_point(c(146.5,-39.1)),crs=4283)
plot(bounds, col = "white")
plot(Victoria_GDA94_GCS$geometry, add = TRUE)
plot(rngPoly$geom, col = scales::alpha(gray(0.5),0.5), add = TRUE)
title(main = "Central Victoria ", adj = 1, font.main = 1, line = -1)
box()
#dev.off()
```
  
Fig. 2.  


```{r fig_3scm_stats, fig.width = 6, fig.height = 5, echo = FALSE}
# Assumed mean annual rainfall for EB calculations = 956.41  - i.e. sum(Croydon$daily$rain_mm)
# Actual mean annual rainfall over 20 year study period was 929 (including the milennium drought)
# For purposes of this exercise, let's use the EB calculator's assumption
# These are the proportion of rainfall values from Zhang (2001)/Walsh et al 2012
# Note that only 79.1% of rainfall becomes impervious runoff - 
#   sum(Croydon$daily$runoff_mm)/sum(Croydon$daily$rain_mm)
Zhang.forest.ro <- 1 - (1 + 2820/956.41)/(1 + 2820/956.41 + 956.41/1410)
Zhang.pasture.ro <- 1 - (1 + 550/956.41)/(1 + 550/956.41 + 956.41/1100)
# As a proportion of impervious runoff, they are:
Zhang.forest.filt.prop <- Zhang.forest.ro / 0.791
Zhang.pasture.filt.prop <- Zhang.pasture.ro / 0.791
scm_per_eia <- cat_stats$n_scm_projects/cat_stats$eia_s1_ha

#pdf("images/Fig 3.pdf", width = 6, height = 5)
# tiff("images/Fig 3.tiff", width = 6*dpi, height = 5*dpi, res = 600, compression = "lzw")
layout(matrix(c(1:5,5), 3,2, byrow = TRUE), heights = c(10,12,1), widths = c(12,12))
par(mar = c(1,4,1,1))
barplot(scm_per_eia , ylim = c(0,20), las = 1, 
        col = c("white",gray(0.75),rep("white",3),gray(0.75)), 
        ylab = "", cex.axis = 1.25, cex.names=1.25)
title(ylab = "No. SCM projects/ ha of EIA", cex.lab = 1.25, line = 2.5)
title(main = "A.", adj = 0, line = 0)
barplot(cat_stats$ia_treated_perc, ylim = c(0,100), las = 1, 
        col = c("white",gray(0.75),rep("white",3),gray(0.75)), 
        ylab = "", cex.axis = 1.25, cex.names=1.25)
title(ylab = "% EIA treated by an SCM", cex.lab = 1.25, line = 2.5)
title(main = "B.", adj = 0, line = 0)
par(mar = c(2,4,1,1))
barplot(cat_stats$vol_filtered_perc, ylim = c(0,100), las = 1, names.arg = cat_stats$cat, 
        col = c("white",gray(0.75),rep("white",3),gray(0.75)), 
        ylab = "", cex.axis = 1.25, cex.names=1.25)
title(ylab = "% Runoff volume filtered", cex.lab = 1.25, line = 2.5)
polygon(c(-1,8,8,-1), 100*c(rep(Zhang.forest.filt.prop,2),rep(Zhang.pasture.filt.prop,2)), 
        col = scales::alpha("red",0.25), border = scales::alpha("red",0.25))
title(main = "C.", adj = 0, line = 0)
barplot(cat_stats$vol_reduced_perc , ylim = c(0,100), las = 1, names.arg = cat_stats$cat, 
        col = c("white",gray(0.75),rep("white",3),gray(0.75)), 
        ylab = "", cex.axis = 1.25, cex.names=1.25)
title(ylab = "% Runoff volume harvested/lost", cex.lab = 1.25, line = 2.5)
polygon(c(-1,8,8,-1), 100*c(1- rep(Zhang.forest.filt.prop,2),1- rep(Zhang.pasture.filt.prop,2)), 
        col = scales::alpha("red",0.25), border = scales::alpha("red",0.25))
title(main = "D.", adj = 0, line = 0)
par(mar = c(0,0,0,0))
plot.new()
title(xlab = "        Sub-catchment", line = -1.25, cex.lab = 1.25)
# dev.off()
```
Fig. 3.  


```{r fig4_MonitoringSiteTrendsA, results="hide", warning = FALSE, echo = FALSE, fig.width = 7, fig.height = 4}
# pdf("images/Fig 4.pdf", width = 7, height = 4)
# tiff("images/Fig 4.tiff", width = 7*dpi, height = 4*dpi, res = 600, compression = "lzw")
layout(matrix(c(1,2,3,4,1,5,6,7,0,8,8,8),3,4,byrow = TRUE),heights= c(10,10,1),widths = c(1,10,10,10))
par(mar = c(0,0,0,0))
plot.new()
title(ylab = expression(paste("Effective imperviousness variant (%)"), sep = ""), line = -1.5, cex.lab = 1.25)
par(mar = c(2,3,1,1))
plotlEItrends(ei_101$iats, xlab = FALSE, cex.axis = 1)
title(main = "A. D4", adj = 0, cex.main = 0.9)
plotlEItrends(ei_103$iats, xlab = FALSE, ylab = FALSE)
title(main = "B. D8", adj = 0, cex.main = 0.9)
plotlEItrends(ei_53$iats, xlab = FALSE, ylab = FALSE)
title(main = "C. Ln", adj = 0, cex.main = 0.9)
plotlEItrends(ei_74$iats, cex.axis = 1)
title(main = "D. L1", adj = 0, cex.main = 0.9) # , font.main = 1, cex.main = 0.9)
legend("bottomright",legend = c(expression(EI[S1]),expression(EI[S]),
                                expression(EI[SF]),expression(EI[SR]),
                                expression(EI[SV]),expression(EI[SW]),expression(EI[S0])),
       lty = c(2,1,1,1,1,1,3), col = c("black",RColorBrewer::brewer.pal(7,"Set1")[c(1:5,7)]), lwd = c(2,1,1,1,1,1,1))
plotlEItrends(ei_36$iats, ylab = FALSE, cex.axis = 1)
title(main = "E. Ls", adj = 0, cex.main = 0.9)
plotlEItrends(ei_71$iats, ylab = FALSE, cex.axis = 1)
title(main = "F. L4", adj = 0, cex.main = 0.9)
par(mar = c(0,0,0,0))
plot.new()
title(xlab = "Date", line = -1.2, cex.lab = 1.25)
# dev.off()
#Explaining odd behaviour in Ln.  
#1. Final drop in EIsw was installation of 4 raingardens along Wattle Valley Rd, treating 1323 m2 of IA reducing EI by ~0.09% by capturing runoff from previously untreated roads. (2016-12-09)
#2. Second big drop in EIsw (2013-11-01) (soon after first big drop which was the initiation of the Morrisons harvesting system on 2013-04-15), with a big drop in EIsf, but not much change in other sub-indices was installation of the Jellyfish system (https://oceanprotect.com.au/jellyfish/) at the end of Fernhill Road.
#3. The fluctuations in EIs, EIsf, EIsr and EIsv post-2015 result from the King Street Upper raingarden being turned on an off. This raingarden diverted flows from Fernhill Road that would otherwise flow to the Jellyfish.  It resulted in changes to catchment relationships, and little change to ultimate performance!
```
  
Fig. 4.  

```{r fig5, echo=FALSE, fig.width = 7, fig.height = 2.5}
ei_ts$lei <- log10(ei_ts$ei*100 + 0.1)
ei_ts$lei_s <- ei_ts$lei_s0 <- ei_ts$lei
exp_cats <- unique(ei_ts_all$sitecode)
for(i in 1:length(exp_cats)){
  ei_ts$lei_s[ei_ts$sitecode == exp_cats[i]] <- log10(ei_ts_all$eb[ei_ts_all$sitecode == exp_cats[i]]*100 + 0.1)
  ei_ts$lei_s0[ei_ts$sitecode == exp_cats[i]] <- log10(ei_ts_all$s[ei_ts_all$sitecode == exp_cats[i]]*100 + 0.1)
}
ei_ts$del_ei_s <-  ei_ts$lei_s - ei_ts$lei
ei_ts$del_ei_s0 <- ei_ts$lei_s0 -ei_ts$lei

line_cols <- siteMap11$col
line_cols[match(c("LIS0001","DBS0008","DBS0004","LIS0004","LSS0001","LSN0001"), siteMap11$sitecode)] <-  
  RColorBrewer::brewer.pal(7,"YlOrRd")[2:7]
 
# pdf("images/Fig 5.pdf", width = 7, height = 2.5)
# tiff("images/Fig 5.tiff", width = 7*dpi, height = 2.5*dpi, res = 600, compression = "lzw")
layout(matrix(c(1,2,3,4,4,4), ncol=3, byrow=TRUE), heights = c(10,1), widths = c(10,10,10))
par(mar=c(2,4,1,1))

 plot(ei_ts$date, ei_ts$lei, type = 'n',
      axes = FALSE, ylab = "", 
      ylim = c(-1,1.5))
 axis(2,at = seq(-2,2,0.5),las = 1)
 axis(1, at = c(0,1e6))
 axis.Date(1, ei_ts$date,las = 1)
 title(ylab = expression(log[10](EI[S1] + 0.1)), line = 2.5)
 for(j in 1:length(siteMap11$sitecode)){
 lines(ei_ts$date[ei_ts$sitecode == siteMap11$sitecode[j]], 
       ei_ts$lei[ei_ts$sitecode == siteMap11$sitecode[j]], type = 'l', col = line_cols[j])
 }
 title("A.",adj = 0)
 
 plot(ei_ts$date, ei_ts$del_ei_s, type = 'n', 
      axes = FALSE, ylab = expression(~Delta~EI[S]), ylim = c(-2,0.5))
 axis(2,at = seq(-3,2,0.5),las = 1)
 axis(1, at = c(0,1e6))
 axis.Date(1, ei_ts$date,las = 1)
 for(j in 1:length(siteMap11$sitecode)){
 lines(ei_ts$date[ei_ts$sitecode == siteMap11$sitecode[j]], 
       ei_ts$del_ei_s[ei_ts$sitecode == siteMap11$sitecode[j]], type = 'l', col = line_cols[j])
 }
 title("B.",adj = 0)
 
 plot(ei_ts$date, ei_ts$del_ei_s0, type = 'n', 
      axes = FALSE, ylab = expression(~Delta~EI[S0]), ylim = c(-2,0.5))
 axis(2,at = seq(-3,2,0.5),las = 1)
 axis(1, at = c(0,1e6))
 axis.Date(1, ei_ts$date,las = 1)
 for(j in 1:length(siteMap11$sitecode)){
 lines(ei_ts$date[ei_ts$sitecode == siteMap11$sitecode[j]], 
       ei_ts$del_ei_s0[ei_ts$sitecode == siteMap11$sitecode[j]], type = 'l', col = line_cols[j])
 }
 title("C.",adj = 0)
 
 legend("bottomleft", lty = 1, 
        col = c(siteMap11$col[siteMap11$sitecode == "BRS0015"], 
                siteMap11$col[siteMap11$sitecode == "LYR0007"],
                RColorBrewer::brewer.pal(7,"YlOrRd")[2:7]),
        legend = c("C","R","L1","D8",
                   "D4","L4","Ls","Ln"),
        cex = 0.9)

 par(mar = c(0,0,0,0)); plot.new(); title(xlab = "      Date", line = -1.5)
# dev.off()
```
Fig. 5

```{r fig6, fig.width = 7, fig.height = 5}
# Code for producing Fig 6 is the final chunk in 'Walsh_et_al_foundation_PLOSW_S5.Rmd'
# The code requires compiling of the Stan model objects (earlier chunks in that
# document), which takes ~ 1 h. Hence printed as a pdf here.
knitr::include_graphics("images/Fig 6.pdf")
```
Fig. 6

##### page break   

## Reference  


