---
title: 'Linking stormwater control performance to stream ecosystem outcomes: incorporating a performance metric into effective imperviousness'
subtitle: 'S5 Appendix. Statistical model for assessing stream responses to stormwater control measures'
author: 'Christopher J. Walsh, Matthew J. Burns, Tim D. Fletcher, Darren G. Bos, Peter Poelsma, Joshphar Kunapo and Sam J. Imberger'
date: 'School of Ecosystem and Forest Sciences, The University of Melbourne, 500 Yarra Boulevard, Burnley Victoria 3121, Australia '
output: 
  word_document:
    reference_docx: officedown_template.docx
csl: plos.csl
bibliography: references.bib
editor_options: 
  chunk_output_type: inline
---

```{r  load_data, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf", message=FALSE) 
dpi <- 600 
requiredPackages <- c("rstan","loo","coda","sjPlot")
lapply(requiredPackages, require, character.only = TRUE)
rstan_options(auto_write = TRUE)
options(mc.cores = 4)
source("load_ld_scms_tables.R")
current.seed <- 787603905 #
set.seed(current.seed)  #to get consistent, repeatable results of analyses
```

### Introduction

In this document, we describe a general statistic model for assessing in-stream response in stream reaches (sites) to catchment-scale stormwater control in urbanized catchments using a Before-After-Reference-Control-Impact experimental design, such as described in this paper, where the after period is characterised by a protracted period of increased stormwater control (in the 'impact' sites, which we term 'experimental' sites, but not in the control or reference sites), and the degree of stormwater control varies widely among the 'impact' sites.

### Model structure

The model contains two main effects that are used to assess changes in the experimental sites that differ from changes in those sites before implementation of stormwater control and from changes over the entire study period in control and reference sites:

-   *degrd*: degradation caused by conventional urban stormwater drainage, inferred from response to variation in *EI~S1~*(effective imperviousness not accounting for stormwater control measures, SCMs) between catchments. Variation in *EI~S1~*is primarily spatial, ranging from near-zero in reference streams to \~25% in the most urban catchment at the end of the study. Growth in *EI~S1~*in each catchment was small compared to differences between sites.

-   *restr*: restoration resulting from reduced stormwater impacts resulting from SCMs, inferred from responses over time to decreasing *EI~S~* (log~10~(*EI~S~*+0.1) - log~10~(*EI~S1~*+0.1)), where *EI~S~* is effective imperviousness accounting for SCM performance: see main text and S1). *EI~S~*is nearzero in all sites except experimental sites, in which it becomes increasingly negative with increasing SCM implementation. 

We modelled each simulated response variable as being drawn from a normal distribution thus:

$$
y_i \sim  Normal(\mu_i, \sigma)
$$ where $y_i$ is the response variable in the *i*th of 76 equally spaced samples, $\mu_i$ is the mean estimate for that sample, and $\sigma$ is the standard deviation of the estimate. We modeled $\mu_i$ in response to predictors as a hierarchical linear model:

$$
\mu_i = \alpha + \alpha [site_j] + \beta_D degrd_i + \beta_R restr_i + \beta_{T}[site_j] t_i + \beta_A autoT_i
$$

where $\alpha$ is the global intercept; $\alpha [site_j]$ is the random variation to that intercept for site *j*; $\beta_D$ represents the effect of *degrd*; $\beta_R$ represents the effect of *restr* (only non-zero in experimental sites after SCM manipulation had begun); $\beta_{T}[site_j]$ represents the effect of time, t, within the site *j*; $\beta_A$ represents the effect of temporal autocorrelation, *autoT*, among samples from each site; a. To derive the temporal autocorrelation variable, we first derived the model without an *autoT* term, and used the residual value from that model for the preceding time in each site.

### Simulated response variables

We simulated contrasting response variables to illustrate the model's ability to detect change in variables with differing responses to stormwater control. We used a simplified version of our study's data. To remove the complication of L4 and D8 being downstream of other sites (requiring introduction of a spatial autocorrelation term), we constructed a data set of predictor variables for the 9 other sites, which are independent of each other.

The simulated data set was populated with four samples per year from 2001 to 2019, from each of the 9 sites, and each date was attributed the observed values of *degrd* (*EI~S1~*) and *restr* ( $\Delta$*EI~S~*). Strictly, *EI~S1~* is effective imperviousness without considering stormwater control measures (SCMs), but for analysis of the experiment, we calculated *EI~S1~* by giving informally drained impervious surfaces a weighting of 0 (as is usual for calculation of EI), and giving impervious surfaces draining to SCMs that were in place before the experiment a weighting of *S* of those SCMs. By doing this, we ensure that *restr* of all sites is 0 before the implementation of the first experimental SCMs[^1]. The *restr* effect is thus only assessing the effect of the experimental SCMs, and the (small) effect of the few existing SCMs becomes part of the *degrd* effect. Arguably, the informally drained impervious surfaces not counted in *EI~S1~* are being intercepted and treated by the natural stormwater control of catchment soils, and the calculation is assuming that treatment is perfect (*S* = 0).

[^1]: This adjustment made no discernable difference to the results. The same analysis with *EI~S1~* calculated without considering existing SCMs produced near-identical results.

The following code creates a data frame of 76 sample dates, with values of *EI~S1~* and $\Delta$*EI~S~* for each date, and plots their trends over time for each of the 9 sites in the analysis (Fig. S5-1).

```{r prepare_sim_set, fig.width = 7, fig.height = 3}
par(mar = c(4,4,1,1))
dates <- seq.Date(from = as.Date("2001-02-01"),
                  to = as.Date("2019-11-01"), 
                  by = "3 months")
t <- 1:length(dates)
sitecodes <- siteMap11$sitecode[!siteMap11$sitecode %in% c("LIS0004","DBS0008")]
sitecodes_p <- c("Sa","Ly","Ol","Ls","Ln","Br","D4","Fe","L1")[match(
  c("SAS0002","LYR0007","OLN0009","LSS0001","LSN0001",
    "BRS0015","DBS0004","FER0006","LIS0001"),
  sitecodes)]
# Adjust ei so that del_ei is zero before experiment in all sites
# i.e. include existing SCMs in calculation of ei_s1
exp_cats <- unique(ei_ts_all$sitecode)
for(i in 1:length(exp_cats)){
  if(ei_ts_all$ei[ei_ts_all$sitecode == exp_cats[i]][2] != 
     ei_ts_all$eb[ei_ts_all$sitecode == exp_cats[i]][2]) 
#    cat(exp_cats[i], "\n")
    ei_ts_all$ei[ei_ts_all$sitecode == exp_cats[i]] <- 
      ei_ts_all$ei[ei_ts_all$sitecode == exp_cats[i]] - 
      (ei_ts_all$ei[ei_ts_all$sitecode == exp_cats[i]][2] - 
      ei_ts_all$eb[ei_ts_all$sitecode == exp_cats[i]][2])
    ei_ts_all$s[ei_ts_all$sitecode == exp_cats[i]] <- 
      ei_ts_all$s[ei_ts_all$sitecode == exp_cats[i]] - 
      (ei_ts_all$s[ei_ts_all$sitecode == exp_cats[i]][2] - 
      ei_ts_all$eb[ei_ts_all$sitecode == exp_cats[i]][2])
    ei_ts$ei[ei_ts$sitecode == exp_cats[i]] <- ei_ts_all$ei[ei_ts_all$sitecode == exp_cats[i]]
    ei_ts$s[ei_ts$sitecode == exp_cats[i]] <- ei_ts_all$s[ei_ts_all$sitecode == exp_cats[i]]
}

ei_ts$lei <- log10(ei_ts$ei*100 + 0.1)
ei_ts$lei_s <- ei_ts$lei_s0 <- ei_ts$lei
for(i in 1:length(exp_cats)){
  ei_ts$lei_s[ei_ts$sitecode == exp_cats[i]] <- 
    log10(ei_ts_all$eb[ei_ts_all$sitecode == exp_cats[i]]*100 + 0.1)
  ei_ts$lei_s0[ei_ts$sitecode == exp_cats[i]] <- 
    log10(ei_ts_all$s[ei_ts_all$sitecode == exp_cats[i]]*100 + 0.1)
}
ei_ts$del_ei_s <-  ei_ts$lei_s - ei_ts$lei

sim_set <- data.frame(site = rep(sitecodes, each = length(t)),
                      date = rep(dates, length(sitecodes)),
                      t = rep(t, length(sitecodes)),
                      lei_s1 = NA, del_ei_s = NA)

for(i in 1:length(sitecodes)){
  sim_set$lei_s1[sim_set$site == sitecodes[i]] <- 
    ei_ts$lei[ei_ts$sitecode == sitecodes[i] & ei_ts$date %in% sim_set$date]
  sim_set$del_ei_s[sim_set$site == sitecodes[i]] <- 
    ei_ts$del_ei_s[ei_ts$sitecode == sitecodes[i] & ei_ts$date %in% sim_set$date]
}

par(mar = c(4,4.5,1,1))
layout(matrix(1:3,1,3,byrow=TRUE),widths = c(10,10,2))
plot(jitter(sim_set$t, factor = 2), jitter(sim_set$del_ei_s, factor = 4), 
     col = RColorBrewer::brewer.pal(9,"Set1")[match(sim_set$site, sitecodes)],
     ylab = expression(~Delta~EI[S]), xlab = "time (t)", las = 1)
title(main = "A.", adj = 0)
plot(jitter(sim_set$t, factor = 2), jitter(sim_set$lei_s1, factor = 4), 
     col = RColorBrewer::brewer.pal(9,"Set1")[match(sim_set$site, sitecodes)],
     ylab = expression(log[10](EI[S1] + 0.1)), xlab = "Time (t)", las = 1)
title(main = "B.", adj = 0)
par(mar = c(0,0,0,0))
plot.new()
legend("center",pch = 1, col = RColorBrewer::brewer.pal(9,"Set1"), legend = sitecodes_p, 
       title = "Site", cex = 1)
```

#### Fig. S5-1. Distributions of A. $\Delta$*EI~S~* (*restr*) and B. log~10~(*EI~S1~* + 0.1) (*degrd*) as a function of t, the `r dim(sim_set)[1]` sampling dates in the simulated dataset.

The simulated response variables---*y* and *y1*---were hypothetical versions of log~10~(FRP). Individual samples of *y* and *y1* varied among sites as a function of EI as observed for log~10~(median FRP) (Fig. 1A).

The 19-year study period was characterised by a shift from dry conditions of the milennium drought in the first decade to more average conditions in the second. It is therefore conceivable that in-stream response variables varied over that time in response to increasingly wet conditions. For instance, increasingly wet catchments could have increased the contribution of ground water to stream flows, and this effect may have been greater in more urban streams. If there was such a long-term change in stream variables over the study period, there is potential for collinearity between the long-term change, urban growth captured in *degrd* (Fig. S5-1B), and the linear decrease in *restr* as SCMs were progressively installed in the experiment. To ensure that the model adequately controls for such potential collinearity, we added a linear increase to y and y1 over the study period, which varied between a 0.08% increase per year for Ly (*EI~S1~* = 0) to a 3.9% increase per year for Br (*EI~S1~* = 21%). As a result of this differential increase in *EI~S1~* among catchments, the response of *y*  and *y1* increased from a slope of 0.28 at the start of the study (equivalent to the slope of the linear regression shown in Fig. 1A) to a slope of 0.42 at the end of the study (and of 0.35 in the middle of the study).

We made no further manipulation of variable *y*: this variable was thus comparable to FRP observed in the study, but showed no response to *restr*. In contrast, a further trend was added to *y1*, so that it varied with *restr* equivalently to *degrd* at the start of the study (a slope of 0.28). Thus a good model of the simulated data should predict that both *y* and *y1* are positively correlated with *degrd* (with a coefficient of 0.35, the mean *EI~S1~* effect over the study period), while *y* should be unaffected by *restr* (coefficient of 0) and *y1* should be positively correlated with *restr* (coefficient of 0.28).

The following code creates *y* and *y1* for the 76 samples and plots their relationship with log~10~(*EI~S1~* + 0.1) (*degrd*) and $\Delta$*EI~S~* (*restr*) and plots Fig. S5-2. It also creates a scaled time variable (*t_scaled*), which ranges between -1.5 and 1.5.

```{r, fig.width = 7, fig.height = 6}
lFRP50 <- c(-1.56864E+00, -2.22185E+00, -2.09691E+00, -2.22185E+00, -2.52288E+00, -1.88606E+00, -2.04576E+00, -2.52288E+00, -2.52288E+00, -2.69897E+00, -2.00000E+00, -2.39794E+00, -1.92082E+00, -1.92082E+00) + 3 # + 3 to convert from mg/L to ug/L
 EI <- c(1.95167E-01, 4.26416E-03, 3.62570E-02, 4.19733E-03, 0.00000E+00, 9.52280E-02, 4.63966E-01, 0.00000E+00, 0.00000E+00, 1.16397E-03, 1.03733E-02, 6.65377E-03, 3.82980E-01, 5.47130E-02)
 lei <- log10(EI*100 + 0.1)
mod_frp_ei <- lm(lFRP50 ~ lei)
# 1. vary among sites as a function of EI as observed for log(median FRP)
sim_set$y <- rnorm(dim(sim_set)[1], 
                   summary(mod_frp_ei)$coefficients[1,1],
                   summary(mod_frp_ei)$coefficients[1,2]) +  #distribution of model intercept
             summary(mod_frp_ei)$coefficients[2,1] * sim_set$lei_s1 + #model slope
             rnorm(dim(sim_set)[1], 0,
                   summary(mod_frp_ei)$coefficients[2,2])  #distribution of model slope error
# For the simulation, we allow values of y to fall below the detection limit   

# 2. add a linear increase with time unrelated to the SCM implementation, 
# and make it a greater increase with greater EI...
sim_set$y <- sim_set$y + (sim_set$lei_s1 + 1.05)*sim_set$t/(76*7.5)
# ~0.08% increase per year for LYR - ei = 0% lei_s1 = -1: 10^(0.05*4/(76*7.5))
# ~3.9% increase per year for BRS  - ei = 21.8% lei_s1 = 1.337589: 10^((1.05 + 1.337829)*4/(76*7.5))
y_0 <- range(sim_set$y); ei_0 <- y_0/0.2836; 
b_0 <- diff(y_0)/diff(ei_0)   #0.2836
y_39 <- y_0 + (ei_0+ 1.05)*38.5/(76*7.5); 
b_39 <- diff(y_39)/diff(ei_0) #0.3511
y_76 <- y_0 + (ei_0+ 1.05)*76/(76*7.5); 
b_76 <- diff(y_76)/diff(ei_0) #0.4169
# which increases the effect of ei on y from 0.284 (b_0) when t = 0 to 0.351 (b_39) when t = 38.5 
# (mean t, half-way through the experiment), and 0.42 at end of study (b_76)

# 3. create variable y1 so that its response to del_ei equals its response to ei_s1
sim_set$y1 <- sim_set$y
sim_set$y1 <- sim_set$y + 
              summary(mod_frp_ei)$coefficients[2,1] * sim_set$del_ei_s  #model slope

# 4. create a time variable scaled to range between -1.5 and 1.5
sim_set$t_scaled <- (sim_set$t - 38.5)/25

# 5. # Centre and scale degrd (lei_s1) and restr (del_ei_s) to prevent a small 
     # number of divergent transitions
sim_set$degrd <- 2*(sim_set$lei_s1 - (min(sim_set$lei_s1) + 
                diff(range(sim_set$lei_s1))*0.5) )/ diff(range(sim_set$lei_s1))
sim_set$restr <- 2*(sim_set$del_ei_s - (min(sim_set$del_ei_s) + 
             diff(range(sim_set$del_ei_s))*0.5) )/ diff(range(sim_set$del_ei_s))

# 5. # Prepare set of values for the model to predict to  # ranging between -1 
#      (order of magnitude reduction in EIs) and 0 (no change in EIs)
restr_p <- 2*(c(-1,0) - (min(sim_set$del_ei_s) + 
            diff(range(sim_set$del_ei_s))*0.5) )/ diff(range(sim_set$del_ei_s))
#make relevant restr_p values equal actual achieved reductions
restr_p <- c(restr_p[1],
             min(sim_set$restr[sim_set$site == "LSN0001"]),
             min(sim_set$restr[sim_set$site == "LIS0001"]),
             min(sim_set$restr[sim_set$site == "LSS0001"]),
             min(sim_set$restr[sim_set$site == "DBS0004"]),
             restr_p[2])
# Make predictions for a reference site, a control site and the 4 experimental 
# sites used in sim_set with ei set at level at end of experiment
lei_s1_p <- log10(ei_ts$ei[ei_ts$sitecode %in% c("LYR0007","BRS0015","DBS0004",
                                   "LSN0001","LIS0001","LSS0001") & 
                ei_ts$date == max(ei_ts$date)]*100 + 0.1)
degrd_p <- 2*(lei_s1_p  - (min(sim_set$lei_s1) + 
                diff(range(sim_set$lei_s1))*0.5) )/ diff(range(sim_set$lei_s1))
new_X <- model.matrix(~ degrd_p + restr_p + t_p + site_p,
                      expand.grid(degrd_p = degrd_p,
                                  restr_p = restr_p, 
                                  t_p = max(sim_set$t_scaled),
                                  site_p = 1:9))
#Convert sitecodes to integers (site_no)
sim_set$site <- factor(sim_set$site, 
                       levels = c("SAS0002","LYR0007","OLN0009", "BRS0015", 
                                  "FER0006", "DBS0004", "LSS0001", "LSN0001", "LIS0001"))
sim_set$site_no <- as.numeric(sim_set$site)

layout(matrix(c(1:5,3),2,3,byrow=TRUE),widths = c(10,10,2), heights = c(10,10))
plot(sim_set$lei_s1, sim_set$y, 
     col = RColorBrewer::brewer.pal(9,"Set1")[match(sim_set$site, sitecodes)],
     xlab = expression(log[10](EI[S1] + 0.1)), ylab = "y", las = 1)
title(main = "A.", adj = 0)
plot(sim_set$lei_s1, sim_set$y1, 
     col = RColorBrewer::brewer.pal(9,"Set1")[match(sim_set$site, sitecodes)],
     xlab = expression(log[10](EI[S1] + 0.1)), ylab = "y", las = 1)
title(main = "B.", adj = 0)
par(mar = c(0,0,0,0))
plot.new()
legend("center",pch = 1, col = RColorBrewer::brewer.pal(9,"Set1"), legend = sitecodes_p, 
       title = "Site", cex = 1)
par(mar = c(4,4,1,1))
plot(sim_set$del_ei_s, sim_set$y, 
     col = RColorBrewer::brewer.pal(9,"Set1")[match(sim_set$site, sitecodes)],
     xlab = expression(~Delta~EI[S]), ylab = "y", las = 1)
title(main = "C.", adj = 0)
plot(sim_set$del_ei_s, sim_set$y1, 
     col = RColorBrewer::brewer.pal(9,"Set1")[match(sim_set$site, sitecodes)],
     xlab = expression(~Delta~EI[S]), ylab = "y1", las = 1)
title(main = "D.", adj = 0)
```

#### Fig. S5-2. Distributions of y (A, C) and y1 (B, D) against log~10~(*EI\~S1* + 0.1) (*degrd*) (A, B) and $\Delta$*EI~S~* (*restr*) (C, D).

### Model compilation and sampling

For each model, we estimated the posterior distributions of all parameters using the Hamiltonian Monte Carlo (HMC) sampler implemented in Stan [@CarpenterEtAl_2017]. We called Stan from R [@rcoreteam_2020] using the library rstan [@StanDevelopmentTeam_2020]. All $\alpha$ and $\beta$ parameters were drawn from normal distributions with estimated means.  We specified prior distributions of $\alpha$ and all $\beta$ parameters as diffuse normal distributions (mean 0, standard deviation 5), and of $\alpha$ [*site~j~*] as diffuse normal distributions (mean zero and estimated standard deviation). Prior distributions for standard deviation of $\alpha$ [*site~j~*] and $\sigma$ were half-Cauchy distributions (mean 0, scale 2.5). We drew inference from 10,800-12,800 posterior samples taken from 4 unthinned chains (2,700 per chain without *autoT* and 3,200 with *autoT,* having discarded 1,500 warm-up values of each chain).

We diagnosed convergence by visually inspecting the HMC chains for adequate mixing and stationarity and using the Gelman-Rubin statistic [with values $\hat R$ \< 1.1 indicating convergence; @GelmanEtAl_2004]. We also ensured other standard diagnostic tests for effective sample size, Bayesian Fraction of Missing Information, and saturation of tree depth [@StanDevelopmentTeam_2021] were satisfied. We quantified model accuracy by calculating the correlation coefficient between predicted and observed (simulated) values of y and y1 for each of the models with an *autoT* term. Most importantly, we assessed if the models accurately detected the simulated *degrd* and *restr* effects for the variables y and y1: (degrd coefficient 0.35 for both variables, and *restr* coefficient 0.28 for y1 and 0 for y).

### Model evaluation

The two models predicted the observed (simulated data) well (Fig.S5-3, *R* = 0.97 for both models), affording confidence in the models' estimates of treatment effects.

```{r fig.width = 6.5, fig.height = 3.25, echo = FALSE}
par(mfrow = c(1,2))
load("~/Documents/git/lsc_dbs_scms/compiled_models/sim_modfit_autot_y1.rda")
# auto_t not calculated for first record from each site, so remove these from prediction
d <- sim_set
d$auto_t <- NA
for (j in 1:length(sitecodes)) {
  d$auto_t[d$site == sitecodes[j]][-1] <- 0
}
d <- d[!is.na(d$auto_t),]
y_pred <- as.data.frame(sim_set_mod, include = TRUE, pars = 'y_pred')
d$y_pred_mean <- apply(y_pred, 2, mean)
plot(d$y1, d$y_pred_mean, pch = 19, col = alpha("black",0.2), axes = FALSE,
     xlab = "Observed (simulated) y1", ylab = "Predicted y1")
axis(1);axis(2,las=1);box(bty='l')
abline(lm(y_pred_mean ~ y1, data = d), col = "red")
abline(0,1,lty = 3)
title(main = "A.", adj = 0)
y_r <- cor(d$y_pred_mean, d$y1)
load("~/Documents/git/lsc_dbs_scms/compiled_models/sim_modfit_autot_y.rda")
y_pred <- as.data.frame(sim_set_mod, include = TRUE, pars = 'y_pred')
d$y_pred_mean <- apply(y_pred, 2, mean)
plot(d$y, d$y_pred_mean, pch = 19, col = alpha("black",0.2), axes = FALSE,
     xlab = "Observed (simulated) y", ylab = "Predicted y")
axis(1);axis(2,las=1);box(bty='l')
abline(lm(y_pred_mean ~ y1, data = d), col = "red")
abline(0,1,lty = 3)
title(main = "B.", adj = 0)
y1_r <- cor(d$y_pred_mean, d$y)
```

#### Fig. S5-3. Predicted vs observed values for the models of the simulated variables A. *y1*, and b. *y*. The data points are opaque to show the concentration of overlapping points. The red line in each panel is the line of best fit, and the dotted line is the 1:1 relationship.

The estimated *degrd* effect was 0.35 (95% credible intervals 0.34, 0.36) in both models, equivalent to the known mean effect in the simulated data

The estimates for the *restr* effects matched the known effects for the simulated data: for *y*, the mean estimated *restr* coefficient was 0.01 (95% credible intervals -0.05, 0.07), which did not differ from the simulated *restr* effect of 0, and for *y1*, the mean estimated *restr* coefficient was 0.29 (95% credible intervals 0.24, 0.36), which did not differ from the simulated *restr* effect of 0.28 (Fig. 6 in main paper).

The experimental design and model structure thus permit accurate detection of changes in response to restr in experimental sites, even in the context of varying temporal trends among the study sites.

### Model code

The code below was used to load the data, and run, check, and evaluate the four models (with and without an *autoT* term for *y* and *y1*). The data, and this document in Rmarkdown format, are available in the github repository <https://github.com/cjbwalsh/lsc_dbs_scms>, linked to the Open Science Framework repository [@walsh_etal_2021]: <https://osf.io/57azq/>. Note that the models took \~30 min to run. The following two code chunks are the Stan models, which were saved as model objects for sampling in the following chunk

The model `sim_vary_t`, (without *autoT*) was compiled with the following Stan code, and saved for sampling, below.

```{stan output.var = "sim_vary_t", eval = FALSE}
data {
int<lower=1> N;     // number of data points
// int<lower=1> N2;    // the number of rows in model prediction data set
int<lower=1> G;     // the number of sites (grouping variable)
real y[N];          // response
real degrd[N];      // EI predictor
real restr[N];      // delta-EI predictor
real t[N];          // time predictor
int<lower=1, upper=G> site[N]; // site predictor
// real degrd_p[N2];   // EI predictor for prediction data set
// real restr_p[N2];   // delta-EI predictor for prediction data set
// real t_p[N2];       // time predictor for prediction data set
// int<lower=1, upper=G> site_p[N2]; // site predictor for prediction data set
}

parameters {
real a;          // global intercept
vector[G] a_site;              // the random site adjustment to intercept
real<lower=0> a_site_sigma;    // sd of random site effect
real bd;                 // the degrd (EIs1) parameter
real br;             // the restr (del_EI) parameter
vector[G] bt;           // the t (time) parameter
real<lower=0> sigma;        // the residual model sd
}

 model { 
  vector[N] mu;
  sigma ~ cauchy(0,2.5);
  a_site_sigma ~ cauchy(0,2.5);
  a ~ normal(0,5);
  bd ~ normal(0,5);
  br ~ normal(0,5);

  for(i in 1:G){
   a_site[i] ~ normal(0,a_site_sigma); 
   bt[i] ~ normal(0,5);
  }

  for (i in 1:N)  {
    mu[i] = a + a_site[site[i]] + bd * degrd[i] + br * restr[i] +  bt[site[i]] * t[i];
}
    y ~ normal( mu , sigma );
}

generated quantities {
// vector[N2] new_y_pred; // predicted values on evenly spaced major predictor variables
vector[N] y_pred;      // predicted values for evaluating fit
vector[N] log_lik;     // log-likelihood for calculating WAIC and loo
// for(i in 1:N2){
//   new_y_pred[i] = a + a_site[site_p[i]] + bd * degrd_p[i] + br * restr_p[i] +  bt[site_p[i]] * t_p[i];
//   }
for(i in 1:N){
  y_pred[i] = a + a_site[site[i]] + bd * degrd[i] + br * restr[i] +  bt[site[i]] * t[i];
   }
for (i in 1:N) log_lik[i] = normal_lpdf(y[i] | y_pred[i] , sigma);
}

```

```{r eval = FALSE}
save(sim_vary_t, file = "compiled_models/sim_vary_t.rda")
```

The model `sim_vary_t_autot`, (with *autoT*) was compiled with the following Stan code, and saved for sampling, below.

```{stan output.var = "sim_vary_t_autot", eval = FALSE}
data {
int<lower=1> N;     // number of data points
int<lower=1> N2;    // the number of rows in model prediction data set
int<lower=1> G;     // the number of sites (grouping variable)
real y[N];          // response
real degrd[N];      // EI predictor
real restr[N];      // delta-EI predictor
real t[N];          // time predictor
real autot[N];      // temporal autocorrelation predictor
int<lower=1, upper=G> site[N]; // site predictor
real degrd_p[N2];   // EI predictor for prediction data set
real restr_p[N2];   // delta-EI predictor for prediction data set
real t_p[N2];       // time predictor for prediction data set
int<lower=1, upper=G> site_p[N2]; // site predictor for prediction data set
}

parameters {
real a;          // global intercept
vector[G] a_site;              // the random site adjustment to intercept
real<lower=0> a_site_sigma;    // sd of random site effect
real bd;                 // the degrd (EIs1) parameter
real br;             // the restr (del_EI) parameter
vector[G] bt;           // the t (time) parameter
real bat;           // the temporal autocorrelation parameter
real<lower=0> sigma;        // the residual model sd
}

 model { 
  vector[N] mu;
  sigma ~ cauchy(0,2.5);
  a_site_sigma ~ cauchy(0,2.5);
  a ~ normal(0,5);
  bd ~ normal(0,5);
  br ~ normal(0,5);
  bat ~ normal(0,5);

  for(i in 1:G){
   a_site[i] ~ normal(0,a_site_sigma); 
   bt[i] ~ normal(0,5);
  }

  for (i in 1:N)  {
    mu[i] = a + a_site[site[i]] + bd * degrd[i] + br * restr[i] +  
            bt[site[i]] * t[i] + bat * autot[i];
}
    y ~ normal( mu , sigma );
}

generated quantities {
vector[N2] new_y_pred; // predicted values on evenly spaced major predictor variables
vector[N] y_pred;      // predicted values for evaluating fit
vector[N] log_lik;     // log-likelihood for calculating WAIC and loo
for(i in 1:N2){
   new_y_pred[i] = a + a_site[site_p[i]] + bd * degrd_p[i] + br * restr_p[i] +  
                   bt[site_p[i]] * t_p[i];
   }
for(i in 1:N){
  y_pred[i] = a + a_site[site[i]] + bd * degrd[i] + br * restr[i] +  
              bt[site[i]] * t[i] + bat * autot[i];
   }
for (i in 1:N) log_lik[i] = normal_lpdf(y[i] | y_pred[i] , sigma);
}

```

```{r eval = FALSE}
save(sim_vary_t_autot, file = "compiled_models/sim_vary_t_autot.rda")
```

For each of *y1* and *y*, the following code samples `sim_vary_t`, calculates the residual error for each predicted data point, and uses those residuals to calculate the *autoT* predictor required to sample the final model `sim_vary_t_autot`. It saves each model fit for plotting Fig. 6 in the main paper.

```{r eval = FALSE}
nChains <- 4
nIter <- 4200
nWarmup <- 1500

for (i in 1:2) {#Full loop takes ~ 30 min to run, including saving the model objects
  if (i == 1) {
    y <- sim_set$y1
  }else{
    y <- sim_set$y
  }
data_list <- list(N = dim(sim_set)[1],
                  N2 = dim(new_X)[1],
                  G = length(sitecodes),
                  y = y,
                  degrd = sim_set$degrd,
                  restr = sim_set$restr,
                  t = sim_set$t_scaled,
                  site = as.integer(sim_set$site_no))

load("compiled_models/sim_vary_t.rda")
sim_set_mod <- sampling(sim_vary_t, data = data_list, chains = nChains, iter = nIter,
                    warmup = nWarmup, control = list(adapt_delta = 0.999, max_treedepth = 15),
                    seed = current.seed) # ~ 1 min
sim_set$non_auto_resids <- sim_set$y1 -  
  as.data.frame(summary(sim_set_mod, pars = "y_pred", probs = c(0.5), 
                        use_cache = FALSE)$summary)$`50%`
sim_set$auto_t <- NA
for (j in 1:length(sitecodes)) {
  sim_set$auto_t[sim_set$site == sitecodes[j]][-1] <- 
    sim_set$non_auto_resids[sim_set$site == sitecodes[j]][-sum(sim_set$site == sitecodes[j])]
}
save(sim_set_mod, file = paste0("compiled_models/sim_modfit_",
                                       ifelse(i == 1, "y1", "y"), ".rda"), 
                                       compress = "xz")
rm(sim_set_mod)
gc()
data_list <- list(N = sum(!is.na(sim_set$auto_t)),
                  N2 = dim(new_X)[1],
                  G = length(sitecodes),
                  y = y[!is.na(sim_set$auto_t)],
                  degrd = sim_set$degrd[!is.na(sim_set$auto_t)],
                  restr = sim_set$restr[!is.na(sim_set$auto_t)],
                  t = sim_set$t_scaled[!is.na(sim_set$auto_t)],
                  site = sim_set$site_no[!is.na(sim_set$auto_t)],
                  autot = sim_set$auto_t[!is.na(sim_set$auto_t)],
                  degrd_p = new_X[,2],
                  restr_p = new_X[,3],
                  t_p = new_X[,4],
                  site_p = new_X[,5])

load("compiled_models/sim_vary_t_autot.rda")
sim_set_mod <- sampling(sim_vary_t_autot, data = data_list, chains = nChains, iter = nIter + 500,
                    warmup = nWarmup, control = list(adapt_delta = 0.999, max_treedepth = 15),
                    seed = current.seed) # ~ 1.5, and 9 min to run for y and y1, respectively
save(sim_set_mod, file = paste0("compiled_models/sim_modfit_autot_",
                                       ifelse(i == 1, "y1", "y"), ".rda"), 
                                       compress = "xz")
if (i == 1) remove(sim_set_mod)
}
```

The following code was used for diagnostic checks of each model to confirm all tests were satisfied (results are recorded in comments in the code).

```{r eval = FALSE}
#### Model diagnostics (using, in part, functions from stan_utility.R,
# a utility file available from https://osf.io/8zme2)  ####
if(!"download.OSF.file.R" %in% dir("code"))
download.OSF.file(GUID = "8zme2",file_name = "stan_utility.R", subdir = "code")
source("code/stan_utility.R")
mt_summary <- as.data.frame(summary(sim_set_mod)$summary)
#Rhat for all parameters should be <1.1
max(mt_summary$Rhat)  # y1 model 1.005, y model 1.008 (without auto_t: y1 1.013, y 1.0008)
#Bulk effective sample size should be >100 x number of chains (https://mc-stan.org/misc/warnings.html#Bulk_ESS)
mon <- as.data.frame(monitor(sim_set_mod))
min(mon$Bulk_ESS)/nChains # y1 model 193, y model 177 (without auto_t: y1 131, 125)
row.names(mon[which.min(mon$Bulk_ESS),]) # for parameter a_site_sigma (lp__ for y without auto_t)
#Tail effective sample size, same requirement as for Bulk_ESS
min(mon$Tail_ESS)/nChains # y1 model 137, y model 125 (without auto_t: y1 116, 110), 
row.names(mon)[which.min(mon$Tail_ESS)] #for parameter for parameter a_site_sigma (y1), lp__ (y)
#Tree depth should not have been saturated in any iterations: should be 0%
check_treedepth(sim_set_mod, max_depth = 15)  #y1 model 0%, y models 0%
#Energy Bayesian Fraction of Missing Information should be >= 0.2 in all chains
get_bfmi(sim_set_mod)
# y1 model, 0.4471590 0.4497258 0.3993348 0.3526953
# y model, 0.3877709 0.3280589 0.2893291 0.5001030 (similar without auto_t)
#Number of iterations ending in a divergence should be 0
check_div(sim_set_mod) #  y1 model 0, y model 0 (without auto_t: 1 y1, 0 y)
# inconsequential single divergent transition, given the main model without 
# auto_t had zero divergent transitons.
#Visual checks of traceplots and...conducted using shinystan()
#shinystan::launch_shinystan(sim_set_mod)
```

Code for reproducing Fig. 6 in the main paper.

```{r fig_6, eval=FALSE, fig.width = 7, fig.height = 5}
probs <- c(0.025,0.10,0.5,0.9,0.975)
prob_names <- paste0("q",formatC(probs*1e3, digits = 2, flag = "00",
                                 mode = "integer"))
pars <- c("a","bd","br","bat","bt")
# un-scale and uncentre rest_p and degrd_p values in new_X 
# to make comparable to simulated parameters
new_X_unscaled <- data.frame(new_X)[,-1]
new_X_unscaled$degrd <- new_X_unscaled$degrd * diff(range(sim_set$lei_s1))/2 +  (min(sim_set$lei_s1) + 
                diff(range(sim_set$lei_s1))*0.5)
new_X_unscaled$restr <- new_X_unscaled$restr * diff(range(sim_set$del_ei_s))/2 +  (min(sim_set$del_ei_s) + 
                diff(range(sim_set$del_ei_s))*0.5)
for(i in 1:2){
load(paste0("compiled_models/sim_modfit_autot_y",ifelse(i == 1, "1",""),".rda"))
  coeffs_post <- as.data.frame(sim_set_mod, include = TRUE, pars = pars)
  # un-scale bd and br to make comparable to simulated parameters
  coeffs_post$bd <- coeffs_post$bd * 2/diff(range(sim_set$lei_s1))
  coeffs_post$br <- coeffs_post$br * 2/diff(range(sim_set$del_ei_s))
  coeffs_quantls <- data.frame(par = names(coeffs_post), 
                      q025 = apply(coeffs_post, 2, quantile, probs = probs[1]))
  for(j in 2:length(probs)){
    coeffs_quantls[prob_names[j]] <- apply(coeffs_post, 2, quantile, 
                                           probs = probs[j])
  }
  if(i == 1) {
    coeffs_y1 <- coeffs_quantls
  }else{
      coeffs_y <- coeffs_quantls
      }
  
  new_y_pred_post <- as.data.frame(sim_set_mod, include = TRUE, pars = "new_y_pred")
  new_y_pred_quantls <- data.frame(par = names(new_y_pred_post), 
                  q025 = apply(new_y_pred_post, 2, quantile, probs = probs[1]))
  for(j in 2:length(probs)){
    new_y_pred_quantls[prob_names[j]] <- apply(new_y_pred_post, 2, quantile, 
                                               probs = probs[j])
  }
  if(i == 1){
    new_X_y1 <- cbind(new_y_pred_quantls, as.data.frame(new_X_unscaled)[-1])
  }else{
      new_X_y <- cbind(new_y_pred_quantls, as.data.frame(new_X_unscaled)[-1])
    }
}
new_X_y1$sitecode <- new_X_y$sitecode <- levels(sim_set$site)[match(new_X_y$site_p,1:9)]

tiff("images/Fig 6.tiff", width = 7*dpi, height = 5.5*dpi, res = 600, compression = "lzw")
layout(matrix(c(1,3,2,4,5,5),3,2,byrow = TRUE), 
       heights=c(10,11,2),widths = c(10,10))
par(mar = c(2,4,1,1))
plot(c(-0.1,0.5),c(0.5,2.5), type = 'n', ylab = "", xlab = "", axes = 0)
for(i in 1:2){
  lines(c(coeffs_y$q025[i+1],coeffs_y$q975[i+1]), c(i,i), lwd = 1.5, lend = 1)
  lines(c(coeffs_y$q100[i+1],coeffs_y$q900[i+1]), c(i,i), lwd = 4, lend = 1)
}
points(coeffs_y$q500[2:3], 1:2, pch = 21, bg = "gray", cex = 1.5)
axis(1, at = seq(-1,1,0.1), labels = rep("",21))
axis(2, at = 0:3, labels = c("","degrd","restr",""), las = 1, font = 3)
#c("",expression(EI[S1]), expression(~Delta~EI[S]),"")
abline(v = 0, lty = 3)
title(main = "A. y", adj = 0)
par(mar = c(4,4,1,1))
plot(c(-0.1,0.5),c(0.5,2.5), type = 'n', ylab = "", xlab = "", axes = 0)
for(i in 1:2){
  lines(c(coeffs_y1$q025[i+1],coeffs_y1$q975[i+1]), c(i,i), lwd = 1.5, lend = 1)
  lines(c(coeffs_y1$q100[i+1],coeffs_y1$q900[i+1]), c(i,i), lwd = 4, lend = 1)
}
points(coeffs_y1$q500[2:3], 1:2, pch = 21, bg = "gray", cex = 1.5)
axis(1, at = seq(-1,1,0.1))
axis(2, at = 0:3, labels =  c("","degrd","restr",""), las = 1, font = 3)
#c("",expression(EI[S1]),expression(~Delta~EI[S]),"")
abline(v = 0, lty = 3)
title(main = "C. y1", adj = 0)
title(xlab = "Coefficient")

# Four 4 experimental catchments, predict for del_ei = 0 (no SCMs),
# del_ei = minimum achieved del_ei achieved in that catchment, and 
# del_ei = -1 (if SCMs were to reduce ei by an order of magnitude)
# and also the reference state LYR0007 at end of study period
experimental_cats <- c("LSN0001","LSS0001","DBS0004","LIS0001")
temp <- new_X_y[0,]
for(i in 1:4){
temp <- rbind(temp,
              new_X_y[new_X_y$sitecode == experimental_cats[i] & 
                    new_X_y$degrd == -1 &  # i.e. LYR0007 EI
                    new_X_y$restr == 0,])
temp <- rbind(temp,
              new_X_y[new_X_y$sitecode == experimental_cats[i] & 
              new_X_y$degrd == ei_ts$lei[ei_ts$sitecode == experimental_cats[i] & 
                                               ei_ts$date == max(ei_ts$date)] & 
                round(new_X_y$restr,4) %in% round(c(-1,
               min(sim_set$del_ei_s[sim_set$site == experimental_cats[i]]),0),4),])
  }
temp$x <- rep(c(-0.15,-0.075,0.075,0.15),4) + rep(1:4, each = 4)
temp$col <- rep(RColorBrewer::brewer.pal(5, "RdYlGn")[c(5,4,2,1)],4)

par(mar = c(2,4,1,1))
plot(temp$x, temp$q500, ylim = c(0.5,1.5), pch = 21, bg = temp$col, 
     cex = 1.25, axes = FALSE, ylab = "", xlab = "")
for(i in 1:dim(temp)[1]){
lines(rep(temp$x[i],2),c(temp$q025[i],temp$q975[i]), 
      col = temp$col[i], lwd = 1.5,  lend = 1)
lines(rep(temp$x[i],2),c(temp$q100[i],temp$q900[i]), 
      col = temp$col[i], lwd = 3.5, , lend = 1)
}
points(temp$x, temp$q500, pch = 21, bg = temp$col, cex = 1.25)
axis(2, at = seq(-1,2,0.5), las = 1)
axis(1, at = 0:5, labels = rep("",6))
title("B.", adj = 0)
title(ylab = "y", font.lab = 3)

temp <- new_X_y[0,]
for(i in 1:4){
temp <- rbind(temp,
              new_X_y1[new_X_y1$sitecode == experimental_cats[i] & 
                    new_X_y1$degrd == -1 &  # i.e. LYR0007 EI
                    new_X_y1$restr == 0,])
temp <- rbind(temp,
            new_X_y1[new_X_y1$sitecode == experimental_cats[i] & 
            new_X_y1$degrd == ei_ts$lei[ei_ts$sitecode == experimental_cats[i] & 
                                               ei_ts$date == max(ei_ts$date)] & 
                round(new_X_y1$restr,4) %in% round(c(-1,
                min(sim_set$del_ei_s[sim_set$site == experimental_cats[i]]),0),4),])
  }
temp$x <- rep(c(-0.15,-0.075,0.075,0.15),4) + rep(1:4, each = 4)
temp$col <- rep(RColorBrewer::brewer.pal(5, "RdYlGn")[c(5,4,2,1)],4)

par(mar = c(4,4,1,1))
plot(temp$x, temp$q500, ylim = c(0.5,1.5), pch = 21, bg = temp$col, 
     cex = 1.25, axes = FALSE, ylab = "", xlab = "")
for(i in 1:dim(temp)[1]){
lines(rep(temp$x[i],2),c(temp$q025[i],temp$q975[i]), 
      col = temp$col[i], lwd = 1.5)
lines(rep(temp$x[i],2),c(temp$q100[i],temp$q900[i]), 
      col = temp$col[i], lwd = 3.5)
}
points(temp$x, temp$q500, pch = 21, bg = temp$col, cex = 1.25)
axis(2, at = seq(-1,2,0.5), las = 1)
axis(1, at = 0:5, labels = c("","Ln","Ls","D4","L1",""))
title("D.", adj = 0)
title(xlab = "Experimental site")
title(ylab = "y1", font.lab = 3)
par(mar = c(0,0,0,0))
plot.new()
legend("center", c("No SCMs", "Achieved reduction", 
                   "EIs reduced by factor of 10", "Reference condition"),
       pch = 21, pt.bg = RColorBrewer::brewer.pal(5, "RdYlGn")[c(1,2,4,5)],
       ncol = 4)
dev.off()
```

## References
